--- 
title: "Exploring the Airbnb Market in Singapore"
author: "Jeanette Choong"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
---

# Project Summary

Airbnbs as potential short-term accommodations are practically illegal in Singapore.

Under the 2017 Planning Act, private home owners are permitted to rent our their homes for a minimum of three consecutive months. For HDB flats, home owners may only rent out their homes to non-tourists for a minimum of six consecutive months, under the Housing and Development Act. 

Why are there still so many Airbnb listings in Singapore despite these legal restrictions? 

```{r message=FALSE}
knitr::opts_chunk$set(dpi = 200)
```

```{r message = FALSE}
library(tidyverse)
library(here)
library(sf)
library(tmap)
```

```{r}
listings <- readRDS(here::here("projdata/listings.rds"))
```


```{r}
listings_sf <- st_as_sf(listings, coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(3414)
# convert coordinates to SVY21
```

```{r}
tmap_mode("view")
tm_shape(listings_sf) +
  tm_symbols(col="blue", shape = 21, size = 0.1, border.col = "grey21", border.lwd=0.5)
```

