---
title: "Week 4 Lab Practice"
author: "Jeanette Choong"
date: "2/18/2020"
output: github_document
---

```{r message=FALSE}
library(tidyverse)
```

# 4.2 Sampling resale transactions
```{r message=FALSE}
sales <- readRDS(here::here("data/sales.rds"))

mu <- sales %>% 
  summarize(stat = mean(resale_price))

mu
```

```{r}
sales %>% 
  sample_n(100) %>% 
  summarize(stat = mean(resale_price))
```

```{r}
sample_mean <- function(data, size = 100) {
  data %>% 
    sample_n(size) %>% 
    summarize(stat = mean(resale_price))
}

sampled_means <- rerun(500, sample_mean(sales)) %>% bind_rows()
```

What does the sampled_means table represent?

```{r}
sampled_means %>% 
  ggplot(aes(x = stat)) + geom_histogram() +
  geom_vline(xintercept=mu$stat)
```
# 4.3 Confidence Intervals
```{r message=FALSE}
sales_sample <- sales %>%
  sample_n(500)
```

```{r}
x_bar <- sales_sample %>% 
  summarize(stat = mean(resale_price))

x_bar
```

```{r}
library(infer)
```

```{r}
sales_sample %>% 
  specify(response = resale_price) %>% 
  calculate(stat = "mean")
```

## 4.3.1 Bootstrapping
We can use this same procedure but introduce one more step: generate. This will generate a ‘bootstrapped’ sample. A bootstrap procedure simply takes a sample that is equally sized to the entire dataset (79,100 rows in this case) and it specifcally takes this sample with replacement. This means that we copy one observation from our dataset and add it to the sample but we do not remove this observation from the dataset. We keep doing this until our sample is equal in size to our dataset (see Burt Chapter 10.7 for an additional explanation).

```{r message=FALSE}
bootstrapped_resale_price <-sales_sample %>% 
  specify(response = resale_price) %>% 
  generate(reps = 500) %>% 
  calculate(stat = "mean")
```


```{r}
bootstrapped_resale_price %>% 
  visualise() + 
  geom_vline(xintercept = x_bar$stat)
```



# 4.3.2 Confidence intervals
```{r}
bootstrapped_resale_price %>% 
  get_ci()
```

```{r}
bootstrapped_resale_price %>% 
get_ci(level = 0.98)
```

## 4.3.3 Confidence intervals for a subset
For example, let’s say you have been asked to answer the question: ‘what is the average HDB flat price in Marine Parade?’.

```{r}
x_bar_mp <- sales %>% 
  filter(town == "MARINE PARADE") %>% 
  summarize(stat = mean(resale_price))

x_bar_mp
```

Based on this initial mean, you could answer that the average price is about $540,400. But if you look at the number of transactions this is based on (only 485!) you might feel a bit less confident. Try to use the previous infer methods (specify, generate) to calculate a 95% confidence interval around that mean. Try to answer both through a numerical summary as well as a visualization.

```{r message=FALSE} 
# get bootstrapped sample for MP
bootstrapped_MP<-sales %>% 
  filter(town == "MARINE PARADE") %>% 
  specify(response = resale_price) %>% 
  generate(reps = 500) %>%
  calculate(stat = "mean")
  
```
```{r}
bootstrapped_MP %>% 
  visualise() + geom_vline(xintercept = x_bar$stat)
```
```{r}
bootstrapped_MP %>% 
  get_ci()
```


# 4.4 Comparing means between groups
```{r}
town_means<- sales %>% 
  filter(town == "QUEENSTOWN" | town == "MARINE PARADE") %>% 
  group_by(town) %>% 
  summarise(mean = mean(resale_price))

town_means
```

```{r}
ggplot() + 
  geom_histogram(
    data = sales %>% 
      filter(town == "QUEENSTOWN" | town =="MARINE PARADE"), mapping = aes(x = resale_price, y =..density.., group=town, fill = town),  position = position_dodge())+
  geom_vline(
    data=town_means, 
    mapping=aes(xintercept = mean, group = town, color = town)
  )
```

```{r}
bootstrapped_resale_price_QT <- sales %>% 
  filter(town == "QUEENSTOWN") %>% 
  specify(response = resale_price) %>% 
  generate(reps = 100) %>% 
  calculate(stat = "mean")

bootstrapped_resale_price_QT %>% 
  get_ci()
```


You obtained the confidence interval for the mean resale price of Marine Parade above, in Section 4.3.3. Do the confidence intervals overlap? What do you think overlapping confidence intervals indicate about the significant difference between the two means?
 - note not to assume there is no significant difference just because they overlap. 
 
```{r}
mean_diff <- sales %>% 
  filter(town == "MARINE PARADE" | town == "QUEENSTOWN") %>% 
  specify(formula = resale_price ~ town) %>% 
  calculate(stat = "diff in means", order = c("QUEENSTOWN", "MARINE PARADE"))

mean_diff
```
 
```{r}
null_distribution <- sales %>% 
  filter(town == "MARINE PARADE" | town == "QUEENSTOWN") %>% 
  specify(formula = resale_price ~ town) %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 500, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("QUEENSTOWN", "MARINE PARADE"))
```


```{r}
null_distribution %>% 
  visualise(bins = 100) +
  shade_p_value(obs_stat = mean_diff, direction = "greater")
```

```{r}
null_distribution %>% 
  get_pvalue(obs_stat = mean_diff, direction = "greater")
```

```{r}
sales %>% 
  filter(town == "MARINE PARADE" | town == "QUEENSTOWN") %>% 
  specify(formula = resale_price ~ town) %>% 
  # hypothesize(null = "independence") %>% # disable the null hypothesis
  generate(reps = 500, type = "bootstrap") %>% # set type to bootstrap
  calculate(stat = "diff in means", order = c("QUEENSTOWN", "MARINE PARADE")) %>% 
  get_ci()
```






