---
title: "Session 10"
author: "Jeanette Choong"
date: "4/14/2020"
output: github_document
---

```{r message = FALSE}
library(tidyverse)
library(spdep)
library(tmap)
```


# Spatial autocorrelation

Step 1: Construct a list of neighbours for each area

```{r}
remaining_lease <- readRDS(here::here("data/remaining_lease.rds"))

hex_sp <- as(remaining_lease, 'Spatial')

hex_neighbours <- poly2nb(hex_sp)

plot(hex_neighbours, coordinates(hex_sp))
```

Step 2: Assign a weight to the relationship to each neighbour. i.e. if a hexagon has 4 neighbours, each neighbour will weigh 1/4. This is similar to doing the moving average in space rather than over time. 

```{r}
hex_weights <- nb2listw (hex_neighbours, style="W", zero.policy = TRUE)
```

Step 3: calculate a spatially lagged version of the variable of interest. In this case, this creates an average years of lease left for all the neighbours of a hexagon. ```lag.listw``` is an ```spdep``` function that does just that. 

```{r}
lease_lag <- lag.listw(hex_weights, hex_sp$remaining_lease)
```

```lease_lag``` is a list of length equal to the number of hexagons we have. Before plotting it in, we need to assign ```lease_lag``` value to its hexagon. 

Step 4: Join ```lease_lag``` to  ```remaining_lease``` by the hex_id


```{r}
lease_lag_df <- remaining_lease %>% 
  add_column(lease_lag = lease_lag)

current.mdoe <- tmap_mode("plot")
tm_shape(lease_lag_df) + tm_polygons(col="lease_lag")
```

### Exercise

Compare the plot of ```lease_lag``` that we have just produced with the plot of ```remaining_lease``` (you can reuse ```lease_lag_df``` since it has a ```remaining_lease``` column, or try out tm_facets). What do you observe?


```{r}
tm_shape(lease_lag_df) + tm_polygons(col="remaining_lease")
```

The ```lease_lag``` plot was smoothing over space


## Moran's I

Step 5: Compute spatial autocorrelation - the measure of correlation between the average remaining lease at one point to the average remaining lease in points around it. 

```{r}
ggplot(lease_lag_df) + 
  geom_point(aes(x=remaining_lease, y = lease_lag))
```

What is this plot telling us?

* When ```remaining_lease``` is low, we also see a low spatially lagged variable (```lease_lag```)
* When ```remaining_lease``` is high, we also see a high spatially lagged variable (```lease_lag```)

What this means is the value we measure in one hexagon seems to be positively dependent on the value we measure in neighbouring hexagons.

In other words, there is a positive autocorrelation between ```remaining_lease``` and ```lease_lag```.

To measure this correlation, we compare the ```lease_lag``` with the original ```remaining_lease``` variable by regressing the original variable on the spatially-lagged version. We call this metric the ‘Moran’s I’.

```{r}
moran_manual <- lm(lease_lag ~ remaining_lease, data=lease_lag_df)

moran_manual
```

In social data, a correlation of 0.5 is considered a fairly strong positive correlation. 

Adding the regression line:

```{r}
ggplot(lease_lag_df,
       aes(x = remaining_lease, y = lease_lag)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE)
```


```{r}
moran_auto <- moran.test(lease_lag_df$remaining_lease, hex_weights)
```


## Local value of autocorrelation

Just as we can calculate spatial autocorrelation globally, we can also look at this on a more local scale. 

In ```spdep``` Moran's I has a local counterpart called ```localmoran```. 
Instead of a global metric, it will give an indication of how much clustering is present around each polygon and indicates whether the neighbourhood values indicate a cold’ cluster or a ‘hot’ cluster. In our case: cold means fewer years left on the lease!


```{r}
moran_local <- localmoran(lease_lag_df$remaining_lease, hex_weights)
lease_lag_df$moran <- moran_local[,1] # extracting the first column of moran_local i.e. the local indicator (li)

tm_shape(lease_lag_df) +
  tm_polygons(col = "moran", palette = "-RdBu")
```

local moran takes in the point statistics for remaining lease and the weights.Punggol new flats = hot spot. Outram, old estate = cold spot. 

'li' = local indicator
