---
title: "Week 2"
author: "Jeanette Choong"
date: "2/6/2020"
output: github_document
---

```{r message=FALSE}
library(tidyverse)
library(here)
```


```{r message=FALSE}
sales <- read_csv(here::here("data/hdb_resale_2015_onwards.csv"))
```

```{r}
sales %>%  glimpse()
```
```{r}
library(lubridate)
library(forcats)

sales %>% 
  mutate(month = ymd(month, truncated = 1), 
         flat_type = as_factor(flat_type), 
         storey_range = as_factor(storey_range),
         flat_model = as_factor(flat_model))
```

# What is the earlier lease commencement date and where are these flats located?

1966, Jurong East
```{r}
sales %>% 
  arrange(lease_commence_date) %>% 
  select(town, lease_commence_date)
```

# What are the largest HDB flats in Singapore? How much did they sell for?

3 room terrace flats in Kallang/Whampoa, sold for between S$830,000 - S$1,185,000.
```{r}
sales %>% 
  arrange(-floor_area_sqm) %>% 
    select(town,floor_area_sqm, flat_model, resale_price)
```

# What is the most expensive flat in Punggol?

5-room Premium Apartment Loft at Blk 305B Punggol Rd, between 16-18 storey.
```{r}
sales %>% 
  filter(town == "PUNGGOL") %>% 
  arrange(-resale_price) %>% 
      select(town, floor_area_sqm, flat_model, resale_price)
```

# Which town has, on average, the largest flats (by floor area)?

Pasir Ris

```{r}
sales %>% 
  arrange(-floor_area_sqm) %>% 
  group_by(town) %>% 
  summarise(average_size = mean(floor_area_sqm)) %>% 
arrange(-average_size)
```

# Which town has, on average, the cheapest flats per square meter?

Choa Chu Kang
```{r}
sales %>% 
mutate(cost_sqm = (resale_price/floor_area_sqm)) %>% 
  group_by(town) %>% 
  summarise(cost_sqm_town = mean(cost_sqm)) %>% 
  arrange(cost_sqm_town)
```

